---
title: "Predicting Change"
output:
  html_document:
    code_folding: hide
    df_print: paged
    theme: flatly
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( echo=TRUE, message=FALSE, warning=FALSE, fig.width=10 )
```

```{r, include=FALSE}
# Libraries
library( dplyr )       # data wrangling
library( knitr )
library( stargazer )
library( scales )
library( geojsonio )   # read shapefiles
library( sp )          # work with shapefiles
library( sf )          # work with shapefiles - simple features format
library( mclust )      # cluster analysis 
library( tmap )        # theme maps
library( ggplot2 )     # graphing 
library( ggthemes )    # nice formats for ggplots
library( pander )      # formatting RMD tables
library( tidycensus )
library( cartogram )  # spatial maps w/ tract size bias reduction
library( maptools )   # spatial object manipulation 
library(gridExtra)
library( here )

source("apikey.R")
census_api_key(api_key)

s.type <- "html"
```

# Predicting Change Based on 2000 Neighborhood Characteristics

By analyzing specific 2000 census data, we can predict gentrification within neighborhoods. First we must establish what consists of a neighborhood that can be a candidate for gentrification. The two variables used to identify these candidates were:

- ***Poverty in higher than normal rate:*** By finding tracts showing a poverty rate that lies higher than the 75% percentile for all urban cities, we can say that tract might have potnential for gentrification.

- ***Median income for metros less than normal rate:*** By finding tracts that have a median income less than the median for the urban cities ($38,586), we can speculate that the tract might have potential for gentrification.

Next, we need to establish which factors signify whether a neighborhood has been gentrified. The following variables were used for this process:

- ***Less Uneducated:*** We look for neighborhoods that show a decrease in uneducated people as a signal for gentrification.

- ***Less Poverty:*** We look at neighborhoods that show a decrease in poverty in the ten year span as an indicator for gnetrification.

- ***Less Unemployed:*** We look for areas that show a decrease in unemployment showing possible gentrification.

- ***Less Renters:*** We look at tracts that have a decrease in renters, singifying that people are more likely moving into housing units instead, showing possible gentrification.

- ***More Money:*** Finally, we look at city-wide values for cities that show an increase in income.

Now that we have variables to address both neighborhoods that are candidates for gentrification and neighrbohoods that have been gnetrified, we can start our coding process.

## Load and Merge Data

With those variables in mind, we begin our process of analyzing the data. The first step is to download data for the years 2000 and 2010 along with their metadata and combine them into one dataset.

```{r}
# Load data
data_2000 <- readRDS( here("data/rodeo/LTDB-2000.rds" ))
data_2010 <- readRDS( here("data/rodeo/LTDB-2010.rds" ))
meta_data <- readRDS( here("data/rodeo/LTDB-META-DATA.rds" ))

data_2000 <- select( data_2000, - year )
data_2010 <- select( data_2010, - year )

# Merge 2000 and 2010 data and Metadata
merged_data <- merge( data_2000, data_2010, by="tractid" )
merged_data <- merge( merged_data, meta_data, by="tractid" )
```

## Filter for Urban cities

Since we're interested in urban areas, we filter the data for only urban tracts.

```{r}
# Filter for Urban cities
data_urban <- filter (merged_data, urban == "urban")
```

## Create and Adjust Median Home Value data

Next, we'll make some quick calcuations and adjustments to the Median Home Value data. First, we'll adjust the 2000 Median Home Value for inflation. Next we'll find the change between the two years, following up with the percent change as well. Finally, we add those values back into the dataset.

```{r}
# adjust 2000 home values for inflation 
mhv.00 <- data_urban$mhmval00 * 1.28855  
mhv.10 <- data_urban$mhmval12

# find Median Home Value change from 2000 to 2010
mhv.change <- mhv.10 - mhv.00

# Fliter out home values less than 10k
mhv.00[ mhv.00 < 1000 ] <- NA
pct.change <- 100 * ( mhv.change / mhv.00 )

#Add variables back into dataset
data_urban$mhv.00 <- mhv.00
data_urban$mhv.10 <- mhv.10
data_urban$mhv.change <- mhv.change
data_urban$pct.change <- pct.change
```

## Calculate Statistical Variables

Here's where we calculate new variables of interest based on our daata. We start by creating percentage variables for a variety of data including race, employment, education, and rent. Next, we create metro-wide variables along with percent change of a variable for the ten year span. finally, we clean up the dataset a bit, by keeping only the variables we want to look at.

```{r}
# Calculate new variables
data_urban <- 
  data_urban %>%
  mutate( 
          # 2000 variables
          p.white.00 = 100 * nhwht00 / pop00.x,
          p.black.00 = 100 * nhblk00 / pop00.x,
          p.hisp.00 = 100 * hisp00 / pop00.x, 
          p.asian.00 = 100 * asian00 / pop00.x,
          p.col.edu.00 = 100 * col00 / ag25up00,
          p.prof.00 = 100 * prof00 / empclf00,
          p.unemp.00 = 100 * unemp00 / clf00,
          pov.rate.00 = 100 * npov00 / dpov00,
          p.rent.00 = 100 * rent00 / hu00,

          # 2010 variables
          p.white.10 = 100 * nhwht10 / pop10,
          p.black.10 = 100 * nhblk10 / pop10,
          p.hisp.10 = 100 * hisp10 / pop10, 
          p.asian.10 = 100 * asian10 / pop10,
          p.col.edu.10 = 100 * col12 / ag25up12,
          p.prof.10 = 100 * prof12 / empclf12,
          p.unemp.10 = 100 * unemp12 / clf12,
          pov.rate.10 = 100 * npov12 / dpov12,
          p.rent.10 = 100 * rent10 / hu10 )

# inflation adjust income 
data_urban$hinc00 <- 1.28855 * data_urban$hinc00

# Calculate new variables using "group_by"
data_urban <-
  data_urban %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.pct.00 = ntile( mhv.00, 100 ),
          metro.mhv.pct.10 = ntile( mhv.10, 100 ),
          metro.median.pay.00 = median( hinc00, na.rm=T ),
          metro.median.pay.10 = median( hinc12, na.rm=T ),
          metro.diversity.rank.00 = ntile( (100-p.white.00), 100 ),
          # metro total population 2000
          metro.pop.total.00 = sum( pop00.x, na.rm=T ),
          # metro total population 2010
          metro.pop.total.10 = sum( pop10, na.rm=T ),
          # tract rank in metro area for unemployment in 2000
          metro.unemp.rank.00 = ntile( p.unemp.00, 100 ),
          # tract rank in metro area for unemployment in 2010
          metro.unemp.rank.10 = ntile( p.unemp.10, 100 ) ) %>%
  ungroup() %>%
  mutate( metro.mhv.pctile.change = metro.mhv.pct.10 - metro.mhv.pct.00,
          # metro population growth 2000-2010
          metro.pop.growth =
            ( metro.pop.total.10 - metro.pop.total.00 ) / metro.pop.total.00,
          metro.unemp.pctile.change = metro.unemp.rank.10 - metro.unemp.rank.00,
          pay.change = metro.median.pay.10 - metro.median.pay.00,
          race.change = p.white.10 - p.white.00,
          pov.change = pov.rate.10 - pov.rate.00,
          p.unemp.change = p.unemp.10 - p.unemp.00,
          p.col.edu.change = p.col.edu.10 - p.col.edu.00,
          p.rent.change = p.rent.10 - p.rent.00)

# Keep only variables we need
data_urban <-           
  data_urban %>%
  select( c( "tractid", "cbsa", "cbsaname",
          "mhv.00", "mhv.10", "mhv.change","pct.change",
          "p.white.00", "p.black.00", "p.hisp.00", "p.asian.00", 
          "p.col.edu.00", "p.prof.00",  "p.unemp.00", "pov.rate.00",
          "metro.mhv.pct.00", "metro.median.pay.00", "metro.pop.total.00",
          "metro.unemp.pctile.change", "metro.unemp.rank.00",
          "metro.pop.growth", "metro.mhv.pctile.change", "pay.change", "race.change",
          "metro.diversity.rank.00", "pov.change", "p.unemp.change",
          "p.col.edu.change", "p.rent.change"
          #"metro.mhv.pct.10", "metro.median.pay.10", "metro.pop.total.10",
          #"p.white.10", "p.black.10", "p.hisp.10", "p.asian.10",
          #"p.hs.edu.10", "p.col.edu.10", "p.prof.10",
          #"p.unemp.10", "pov.rate.10", "metro.unemp.rank.10"
          ) ) 

```

**Back up our new Dataset.**

```{r}
# Back up your data just in case
data_master <- data_urban
```

## Create data for Candidates and Gentrified Neighborhoods

Finally, we can analyze which neighborhooods are good candidates for gentrification while at the same time, figure out which of those neighborhoods were gentrified in the ten year span. First, we'll create logic variables signifying whether that neighborhood reached a certain threshold. Those logic variables are none other than the variables detailed above when deciding on factors. With those created, we filter the dataset for candidates and of those candidates who were gentirified.

```{r}
# education status
# unemployment
# poverty
# median income
# % renters

data_urban <- data_master

# proportion of college educated increased by 3 percent
less.uneducated <- data_urban$p.col.edu.change > 3 

# proportion of unemployment decreased by 3 percent
less.unemployed <- data_urban$p.unemp.change > -3 

# proportion of poverty decreased by 3 percent
less.poverty <- data_urban$pov.change > -3 

# proportion of renters decreased
less.renters <- data_urban$p.rent.change < 0

# median income increased
more.money <- data_urban$pay.change > 0

# poverty in higher than normal rate
poor.2000 <- data_urban$pov.rate.00 > 9 

# median income for metros less than normal rate
income.2000 <- data_urban$metro.median.pay.00 < 38586

# Create logic variables that signify what were looking for
poor.and.income <- poor.2000 & income.2000
g.flag <- poor.2000 & income.2000 & less.uneducated & less.poverty & less.unemployed & less.renters & more.money

# Create subsets of the candidates and the gnetrified tracts
data_cand <- data_urban[poor.and.income,]
data_gentrified <- data_urban[g.flag,]

# Add up how many candidates and how many gentirfied tracts
num.candidates <-  sum( poor.and.income, na.rm=T )
num.gentrified <- sum( g.flag, na.rm=T )
```

# Demographics of the Candidate Neighborhoods

Whew, we did it! With our new claculations, let's first look at a summary of the candidates below, giving us the demographics. Looking at the Statistics below, we can see that the demographics for neighborhood candidates of gentirifcation give us a wide variety of values. Some highlights include a low Median Home Value (less than \$100,000) which is emphasized in the metro-wide Median Home Value Percentile (mean is only 38%), a low mean percentage of College Educated (16%) and Professionally Employed people (25%),a low mean Pay Rate (\$34,677), and a high mean Poverty Rate (21%).

```{r, results="asis"}
data_cand <- data.frame(data_cand)
stargazer( data_cand, 
           type=s.type, 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```

# Neighborhood Health Metrics

Now let's look at a statistical summary of the gentrified neighborhoods, focusing on the health metrics we chose to determine whether a neighborhood was gentrified. There was a 8% growth in College Educated people, much larger compared to just a 3% growth for all candidates. Although the mean for percent change in unemployment did not move much (from a 3% within Candidates to a 5% of Gentrified areas), we see that the minimum and maximum has shortenened, giving a smaller range of change. Looking at the change in Poverty, we find it was quite similar to Unemployment, with neglegent change in mean, but a much lower range of percentage in gentrified areas than the candidates. Looking the increase of Rent, we see that the Candidates actually have a positive mean (2%), while the actual gentrified neighborhoods have a negative mean (-4%) which is what we were looking for. Surpringly, when we look at our final health metric, Pay Change, we see that there was not change at all between Candidates and those Gentrified.

```{r, results="asis"}
data_gentrified <- data.frame(data_gentrified)
stargazer( data_gentrified, 
           type=s.type, 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```

Over all, of the 8,773 Candidates, 8% were genetrified based on this model.

```{r}
num.candidates

num.gentrified / num.candidates
```

# Sample Map of Phoenix, AZ

Now let's take a closer look at Phoenix, Arizona using maps. First, we need to load Phoenix's spacial data.

```{r}
# we have phoenix already packaged and on GitHub for easy load: 
github.url <- "https://raw.githubusercontent.com/DS4PS/cpp-529-master/master/data/phx_dorling.geojson"
phx <- geojson_read( x=github.url,  what="sp" )
```

Then we merge our calculated data with the Phoenix data. 

```{r}
data2 <- data_master
# create small dataframe for the merge
df1 <- data.frame( data2 )

# create GEOID that matches GIS format

# create a geoID for merging by tract 
df1$GEOID <- substr( df1$tractid, 6, 18 )  # extract codes
df1$GEOID <- gsub( "-", "", df1$GEOID )    # remove hyphens

phx1 <- merge( phx, df1, by.x="GEOID", by.y="GEOID" )
```

Finally, we print out a few maps to compare. The following maps explore Median Home Value Percent Change, College Educated Percent Change, Unemployment Percent Change, Percent white People Change, Poverty Percent Change, and Rent Percent Change. Looking at the maps side-by-side, we can visually see a trend with the colors. Using the map that shows Median Home Value percent Change as a basis, all the other maps show a similar trend in color scheme. this shows, that our Neighborhood Health Metrics match with Home Growth to a degree.

The distribution of homes with higher values are located in Paradise Valley, Scottsdale, Gilbert, Chandler, and Ahwatukee. The lower priced value homes are located in West Phoenix, Glendale, and Mesa which is to be expected. The largest gains occur in Central Phoenix including Arcadia, Tempe, and North Phoenix. These homes were average in 2000 but had significant gains.

```{r}
phx1 <- spTransform( phx1, CRS("+init=epsg:3395") )

bb <- st_bbox( c( xmin = -12519146, xmax = -12421368, 
                 ymax = 3965924, ymin = 3899074 ), 
               crs = st_crs("+init=epsg:3395")) 

tmap1 <-
  tm_shape( phx1, bbox=bb ) + 
  tm_polygons( col=c("pct.change","p.col.edu.change", 
                     "p.unemp.change","race.change",
                     "pov.change", "p.rent.change"),
               n=6, ncol=2, style="quantile", palette="Spectral" ) +
  tm_layout( "Dorling Cartogram", title.position=c("right","top"), legend.position = c("left", "bottom") )

tmap1
```

-------
<br>

Analysis Created By: Group 1: Ricky Duran, Jason Eastman, & Katie Gentry
For: CPP 528 - Data Sciences for Public Service III (Spring 2020)
Created On: "`r format(Sys.time(), '%B %d, %Y')`"

<br>




